---
title: "Solution - Sheet 03"
author: 
  - "Jaya Bharatam 235291"
  - "Agatha Anna Baby 235293"
  - "Amrutha Manoharan 236892"
  - "Nishant Madan 230370"
date: "4/21/2022"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(nlme)
load("SimulatedTreatmentEffect-3pLL.RData")
```

```{r}
##Part a)
##In 3pLL model the first parameter theta_1=0 for the entire dataset.
model.3pLL.001 <-
  gnls(
    resp ~ (0 + (th4-0) / (1 + (exp((conc - th2) * th3)))),
    data = conc.resp.df,
    params = list(th2 + th3 + th4 ~ 1),
    start = c(1, 2, 100),
    control = gnlsControl(nlsTol = 0.1)
  )
model.3pLL.001
```

The lower asymptote, ${\theta_1}$, is 0 and the upper asymptote, ${\theta_4}$ is 99.081280, which is nearly close to 100.
At ${\theta_2}$=0.959338, we got the half-maximal response which is (0.959338+99.081280)/2 = 50.02. The parameter ${\theta_3}$=2.305889 > 0, which signifies a decreasing profile for increasing concentrations.

```{r}
##Part b)
##In 3pLL model the first parameter theta_11,theta_12,theta_13=0 for the entire dataset.
dum1 <- ifelse(conc.resp.df$treat == "T1", 1, 0)
dum2 <- ifelse(conc.resp.df$treat == "T2", 1, 0)
dum3 <- ifelse(conc.resp.df$treat == "T3", 1, 0)

model.3pLL.002 <- gnls(
  resp ~ th4 * dum3 +
    (0 + (th4-0) / (1 + (exp((conc - th2) * th3)))),
  data = conc.resp.df,
  params = list(th2 + th3 + th4 ~ 1),
  start = c(1, 2, 100),
  control = gnlsControl(nlsTol = 0.1)
)
model.3pLL.002
```

The lower asymptote, ${\theta_1}$, is 0 and the upper asymptote, ${\theta_4}$ is 80.102577, which is actually not really close to 100.
At ${\theta_2}$=1.271901, we got the half-maximal response which is (1.271901+80.102577)/2 = 40.69. The parameter ${\theta_3}$=3.373962 > 0, which signifies a decreasing profile for each increasing concentrations.

```{r}
##Part c)
model.3pLL.003 <- gnls(
  resp ~ (0 + (th4-0) / (1 + (exp((conc - th21) * th31)))) * (dum1) 
    + (0 + (th4-0) / (1 + (exp((conc - th22) * th32 )))) * (dum2),
  data = conc.resp.df,
  params = list(th21 + th22 + th31 + th32 + th4 ~ 1),
  start = c(1, 1, 2, 2, 100),
  control = gnlsControl(nlsTol = 0.1, apVar = TRUE)
)
model.3pLL.003
```

The model use parameter ${\theta_4}$ is shared across all three treatments and, ${\theta_2}$ and ${\theta_3}$ are estimated separately for the first and the second treatment. 
Lower asymptote ${\theta_1}$ is zero.
${\theta_21}$ 0.730, ${\theta_22}$ 1.189 and the upper asymptote is 99.22, which is nearly close to 100.
The parameter ${\theta_{31}}$=2.3477416 > 0 and ${\theta_{32}}$=2.4848554 > 0, a higher value for the slope parameter, demonstrates a steeper decrease of the curve.

```{r}
#Part d)
anova(model.3pLL.001, model.3pLL.002)
```

Both AIC and BIC value increases after adding the treatment 3 in the second model, which further increases p-value but not enough to cross the significance level of 5%, hence we can clearly reject the null hypothesis that both models are equivalent. This indicates that the second model has a significantly better fit and thus, we can difference between all the three treatment groups can be easily seen.
