---
title: "Solution - Sheet 01"
author: 
  - "Jaya Bharatam 235291"
  - "Agatha Anna Baby 235293"
  - "Amrutha Manoharan 236892"
  - "Nishant Madan 230370"
date: "4/7/2022"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Exercise 1: Descriptive analysis of the VPA dataset

```{r library}
library(ggplot2)
load("E:/TUD/04.Semester/toxicology-I/assignments/1/VPAData-Random.Rda")
concentration <-
  c(rep(0, 6), rep(c(25, 150, 350, 450, 550, 800, 1000), each = 3))
concentrations = rep(unique(concentration), each = 10000)
```

(a) Display the entire dataset via boxplots, analogously to the Figure from the lecture (Chapter
1.4, slide 22), i.e. with one boxplot per concentrations and replicate.

```{r}
df <- data.frame(randomVPA)
colnames(df) <- concentration
colors = c(
  rep("red", 6),
  rep("orange", 3),
  rep("lightgreen", 3),
  rep("green", 3),
  rep("lightblue", 3),
  rep("blue", 3),
  rep("purple", 3),
  rep("pink", 3)
)
boxplot(df, col = colors)
```
Each boxplot, shown above is almost the same variability. For each concentrations they have their maximum lies not above 12 and minimum not below 0. They are grouped and colored as for each concentrations.


(b) For each gene, calculate the standard deviation of all three / six replicates corresponding to
the same concentration This yields 8 values for each gene. Visualize the results stratified by
concentration, i.e. summarize the results via a histogram for each concentration separately.

```{r}
sd.by.concentrations <- t(apply(randomVPA, 1, function(x) {
  tapply(x, concentration, sd)
}))

sd.df <- data.frame(sd = c(sd.by.concentrations),
                    concentration = concentrations)

ggplot(sd.df, aes(x = sd)) +
  geom_histogram(bins = 120) +
  facet_wrap(. ~ concentration) +
  theme_bw() +
  xlab("Standard Deviation of the replicates for each concentrations separately") +
  ylab("Count")
```
All the standard error histograms are positively right-skewed. Some of there peaks lies almost at 0.2, while some other have their maximum value at 0.7 and even more to 1 as well. 

(c) For each concentrations separately, find the gene for which the three / six replicates have the
highest standard deviation. Plot the concentrations-response profiles for the resulting eight
genes. Repeat the analysis, now finding the gene for which the three / six replicates have
the lowest standard deviation, for each concentrations separately. Plot the concentrations-response
profiles for the resulting eight genes.

```{r}
se.by.concentration.nocontrol <-
  t(apply(randomVPA[,-(1:6)], 1, function(x) {
    tapply(x, concentration[-(1:6)], function(y)
      sd(y) / sqrt(3))
  }))
se.control <- apply(randomVPA[, 1:6], 1, function(y)
  sd(y) / sqrt(6))
se.by.concentration <-
  cbind(se.control, se.by.concentration.nocontrol)
se.df <- data.frame(se = c(se.by.concentration),
                    concentration = concentrations)
ggplot(se.df, aes(x = se)) +
  geom_histogram(bins = 120) +
  facet_wrap(. ~ concentration) +
  theme_bw() +
  xlab("Standard Error of the replicates for each concentrations separately") +
  ylab("Count")
```
The results compare to b) are almost similar with a compression of the histograms for all positive
concentrations by the factor 1/sqrt(3) nearly equal to 0.58. The histogram for the control is more compressed, by 1/sqrt(6) nearly equal to 0.41, and hence it becomes even more narrower.

(d) For each gene, calculate the mean expression value per concentrations This yields a reduced
dataset with dimensions 500 and 8. Visualize the reduced dataset via boxplots analogously
to (a), but now with only one plot per concentrations

```{r}
mean.per.concentration <- t(apply(randomVPA, 1, function(x) {
  tapply(x, concentration, mean)
}))
boxplot(mean.per.concentration, col = rainbow(length(unique(concentration))))
```
Each concentration(mean.per.concentration) boxplot is nearly the same as the other. Showing above with different colors as for each concentrations respectively.

(e) Determine the number of genes for which the mean gene expression is monotonically
increasing, and the number of genes for which it is monotonically decreasing. Plot the
concentrations-response profiles for all genes with monotonically increasing and monotonically
decreasing profiles. Make sure to indicate both the individual measurements per
concentrations and the mean expression value per concentrations in the plot.

```{r}
monoton.increasing <- apply(mean.per.concentration, 1, function(x) {
  all(diff(x) > 0)
})
sum(monoton.increasing)
```

```{r}
monoton.decreasing <- apply(mean.per.concentration, 1, function(x) {
  all(c(x[1] > x[2], x[2] > x[3], x[3] > x[4], x[4] > x[5], x[5] > x[6], x[6] >
          x[7], x[7] > x[8]))
})
sum(monoton.decreasing)
```

The number of genes with monotonically increasing expression values (19) and the number of genes with
monotonically decreasing expression values (15) is similar and rather low. Thus, the vast majority of genes do not exhibit a monotonic profile.

```{r fig.show="hold", out.width="50%"}
inc.mon <- which(monoton.increasing == TRUE)
for (i in inc.mon) {
  plot(
    concentration,
    randomVPA[i, ],
    pch = 16,
    col = "grey",
    xlab = "concentrations",
    ylab = "Gene expression value",
    main = rownames(randomVPA)[i]
  )
  points(unique(concentration),
         mean.per.concentration[i, ],
         pch = 16,
         col = "black")
}
```

```{r fig.show="hold", out.width="50%"}
dec.mon <- which(monoton.decreasing == TRUE)
for (i in dec.mon) {
  plot(
    concentration,
    randomVPA[i, ],
    pch = 16,
    col = "grey",
    xlab = "concentrations",
    ylab = "Gene expression value",
    main = rownames(randomVPA)[i]
  )
  points(unique(concentration),
         mean.per.concentration[i, ],
         pch = 16,
         col = "black")
}
```

(f) For each gene, calculate the difference between the mean gene expression value of each
positive concentration and the mean gene expression value of the control. Visualize the
results with one histogram per concentration.

```{r}
f.change <- t(apply(mean.per.concentration, 1, function(x) {
  x[2:8] - x[1]
}))
f.change.df <- data.frame(
  FC = c(
    f.change[, 1],
    f.change[, 2],
    f.change[, 3],
    f.change[, 4],
    f.change[, 5],
    f.change[, 6],
    f.change[, 7]
  ),
  concentration = rep(unique(concentration)[-1], each = 10000)
)
ggplot(f.change.df, aes(x = FC)) +
  geom_histogram(bins = 120) +
  facet_wrap(. ~ concentration, scales = "free_y") +
  theme_bw() +
  xlab("Mean difference in gene expression to control") +
  ylab("Count")
```

```{r}
ggplot(f.change.df, aes(x = FC)) +
  geom_histogram(bins = 150) +
  facet_wrap(. ~ concentration, scales = "free_y") +
  theme_bw() +
  xlab("Mean difference in gene expression to control") +
  ylab("Count") + xlim(-2, 2)
```

Major differences is in the range between -2 and 2. With increasing concentrations, the histograms
become slightly broader, proves larger differences in expression values (“fold changes”). Additionally, indicates that here only very few f.change of approximately 0 are observed.
